---
title: "Extending the Theme System"
author: "zztable1_nextgen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Extending the Theme System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  results = 'asis'
)

# Load package
library(devtools)
load_all()
```

```{r theme-css, results='asis', echo=FALSE}
# Generate and inject CSS for theme styling
theme_css <- generate_theme_css()
cat("<style>\n")
cat(theme_css)
cat("\n</style>")
```

# Introduction

The theme system in zztable1_nextgen allows you to customize the appearance of your Table 1 output for different formats (console, LaTeX, HTML). This vignette demonstrates how to create custom themes and extend the theme system for your specific needs.

## Built-in Themes

zztable1_nextgen comes with several pre-configured themes for major medical journals:

```{r list-themes}
available_themes <- list_available_themes()
cat("Available themes:", paste(available_themes, collapse = ", "), "\n")

# Get details on NEJM theme
nejm_theme <- get_theme("nejm")
cat("\nNEJM Theme Details:\n")
cat("- Name:", nejm_theme$name, "\n")
cat("- Decimal places:", nejm_theme$decimal_places, "\n")
cat("- CSS class:", nejm_theme$css_class, "\n")
```

# Creating Custom Themes

## Basic Custom Theme

The simplest way to create a custom theme is using `create_custom_theme()`, which creates a theme based on an existing theme:

```{r custom-theme-basic}
# Create a custom theme starting from NEJM
my_theme <- create_custom_theme(
  name = "MyCustom",
  base_theme = "nejm",
  decimal_places = 2
)

cat("Custom theme created:\n")
cat("- Name:", my_theme$name, "\n")
cat("- Decimal places:", my_theme$decimal_places, "\n")
cat("- CSS class:", my_theme$css_class, "\n")
```

## Advanced Custom Theme

For more control, create a theme from scratch with individual property parameters:

```{r custom-theme-advanced}
# Create a highly customized theme using individual parameters
corporate_theme <- create_custom_theme(
  name = "Corporate",
  base_theme = "console",
  decimal_places = 1,
  font_family = "Arial, sans-serif",
  font_size = "11pt",
  background_color = "#e8f4f8",
  border_color = "#003366"
)

cat("Corporate theme created with custom properties:\n")
str(corporate_theme[c("name", "decimal_places")])
```

## Modifying Existing Themes

You can also modify an existing theme after creation using `customize_theme()`:

```{r customize-theme}
# Modify an existing theme
modified_lancet <- customize_theme(
  "lancet",
  decimal_places = 3,
  css_properties = list(
    font_size = "12pt"
  )
)

cat("Modified Lancet theme:\n")
cat("- Decimal places:", modified_lancet$decimal_places, "\n")
cat("- Font size:", modified_lancet$css_properties$font_size, "\n")
```

# Using Custom Themes

## With table1() Function

Pass your custom theme to the `table1()` function:

```{r example-custom-theme}
# Sample data
data(mtcars)
mtcars$transmission <- factor(ifelse(mtcars$am == 1, "Manual", "Automatic"))
mtcars$engine <- factor(ifelse(mtcars$vs == 1, "V-Engine", "Straight"))

# Create table with custom theme
bp <- table1(
  transmission ~ mpg + hp + wt,
  data = mtcars,
  theme = "nejm",
  pvalue = TRUE
)

# Display the table (format depends on output format)
cat("Table 1: Vehicle Characteristics by Transmission Type\n")
print(bp)
```

## Applying Themes to Existing Blueprints

If you already have a blueprint created, you can apply a theme using `apply_theme()`:

```{r apply-theme}
# Create blueprint without theme
bp_unthemed <- table1(
  transmission ~ mpg + hp,
  data = mtcars
)

# Apply theme afterward
bp_themed <- apply_theme(bp_unthemed, "lancet")

cat("Theme applied to existing blueprint\n")
```

# Theme Properties Reference

## Structure of a Theme Object

A theme is a list with the following key properties:

```{r theme-structure}
# Examine theme structure
nejm_theme <- get_theme("nejm")

cat("NEJM Theme Structure:\n")
cat("- name:", nejm_theme$name, "\n")
cat("- theme_name:", nejm_theme$theme_name, "\n")
cat("- decimal_places:", nejm_theme$decimal_places, "\n")
cat("- css_class:", nejm_theme$css_class, "\n")
cat("- header_separator:", nejm_theme$header_separator, "\n")
cat("\nCSS Properties:\n")
for (prop in names(nejm_theme$css_properties)) {
  cat("  -", prop, ":", nejm_theme$css_properties[[prop]], "\n")
}
```

## CSS Properties

Themes support the following CSS properties:

| Property | Purpose | Example |
|----------|---------|---------|
| `font_family` | Font used for the table | "Arial, sans-serif" |
| `font_size` | Base font size | "11pt" |
| `header_background` | Header row background color | "#f0f0f0" |
| `header_text_color` | Header text color | "black" |
| `row_stripe_color` | Alternating row color | "#fafafa" |
| `border_color` | Table border color | "#cccccc" |
| `padding` | Cell padding | "8px" |

## Dimension Rules

Themes can specify how variables and factors are displayed:

```{r dimension-rules}
# Get dimension rules from a theme
theme <- get_theme("nejm")

if (!is.null(theme$dimension_rules)) {
  cat("Dimension Rules:\n")
  str(theme$dimension_rules)
} else {
  cat("No custom dimension rules defined\n")
}
```

# Advanced: Output Format Handling

## Understanding Format-Specific Behavior

Themes work across three output formats, with format-specific rendering:

### Console Format
- Plain text output for terminal/console display
- Uses ASCII formatting
- Suitable for interactive analysis

### LaTeX Format
- PDF generation for publication
- Uses LaTeX commands and packages
- Best for journal submissions
- Supports advanced formatting (bold, colors, striping)

### HTML Format
- Web-based tables
- CSS styling for appearance
- Interactive browser display

## Creating Format-Specific Customizations

For advanced use cases, you can create a theme with format-specific properties:

```{r format-specific}
# Advanced theme with format-specific properties
publication_theme <- create_custom_theme(
  name = "Publication",
  base_theme = "nejm",
  decimal_places = 1,
  font_family = "Georgia, serif",
  background_color = "#fefcf0"
)

# This theme works across all formats automatically
# Format-specific rendering adjusts appearance
cat("Publication theme created\n")
```

# Best Practices

## Naming Conventions

- Use descriptive names for your themes
- Include organization name if appropriate
- Avoid special characters
- Examples: "MyOrg_Publications", "Clinical_Standard", "Conference_2024"

## Consistency

- Define themes once and reuse them
- Store custom themes in a package or configuration
- Document your theme choices in methods sections

## Accessibility

- Ensure sufficient color contrast in themes
- Use patterns/striping in addition to colors
- Test output in multiple formats (console, PDF, HTML)

## Testing

Always test your custom themes:

```{r test-theme}
# Test theme with different data types
test_data <- mtcars[1:10, ]
test_data$group <- factor(c(rep("A", 5), rep("B", 5)))

bp_test <- table1(
  group ~ mpg + hp + wt + cyl,
  data = test_data,
  theme = "nejm"
)

# Verify appearance
cat("Testing custom theme...\n")
```

# Troubleshooting

## Common Issues

### Theme Not Found
```{r error-example-eval, eval = FALSE}
# This will use console as fallback
my_table <- table1(x ~ y, data = df, theme = "nonexistent")
```

**Solution:** Use `list_available_themes()` to verify theme names

### CSS Not Applied
Ensure you're using the rendered output in the correct format:

```{r css-formats}
# CSS only applies to HTML/LaTeX output
bp <- table1(transmission ~ mpg, data = mtcars, theme = "nejm")

# Console: plain text, no CSS
console_output <- render_console(bp)

# HTML: CSS applied
html_output <- render_html(bp)

# LaTeX: LaTeX formatting applied
latex_output <- render_latex(bp)
```

### Rendering Issues
If theme colors don't appear in PDF:
- Verify LaTeX packages are included in YAML header
- Check PDF viewer supports colors
- Try a simpler theme first

## Getting Help

For issues with custom themes:
1. Verify theme structure with `str(your_theme)`
2. Check that all required fields are present
3. Test with built-in themes first
4. Review examples in the vignette

# Summary

The theme system provides flexible control over Table 1 appearance:

- **Built-in themes** for major journals (NEJM, Lancet, JAMA, BMJ)
- **Custom themes** via `create_custom_theme()`
- **Theme modification** with `customize_theme()`
- **Multi-format support** (console, LaTeX, HTML)
- **CSS properties** for HTML styling
- **Dimension rules** for variable/factor formatting

With these tools, you can create professional tables tailored to your publication requirements while maintaining consistent formatting across formats.

# References

See also:
- `vignette("theming_system")` - Built-in theme demonstrations
- `vignette("zztable1_nextgen_guide")` - Package overview
- `?create_custom_theme` - Function documentation
- `?apply_theme` - Theme application
