---
title: "Quickstart Guide: zztable1"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zztable1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**zztable1** is a next-generation R package for creating publication-quality
summary tables (Table 1) with a flexible blueprint-based architecture. The
package features lazy evaluation, sparse storage, and medical journal themes
(NEJM, Lancet, JAMA).

## Installation

```{r install}
# Install from GitHub
devtools::install_github("rgt47/zztable1")
```

```{r load}
library(zztable1)
```

## Basic Usage

Create a Table 1 using a formula interface:

```{r basic}
# Formula: grouping_variable ~ variables_to_summarize
table1(arm ~ age + sex + bmi, data = trial_data)
```

### Example with mtcars

```{r mtcars-example}
# Create grouping variable
mtcars$transmission <- factor(mtcars$am, labels = c("Automatic", "Manual"))

# Generate Table 1
table1(transmission ~ mpg + hp + wt + cyl, data = mtcars)
```

## Key Features

### Medical Journal Themes

Publication-ready themes matching actual journal formatting:

```{r themes}
# New England Journal of Medicine
table1(arm ~ age + sex, data = data, theme = "nejm")

# The Lancet
table1(arm ~ age + sex, data = data, theme = "lancet")

# JAMA
table1(arm ~ age + sex, data = data, theme = "jama")

# Minimal (default)
table1(arm ~ age + sex, data = data, theme = "minimal")
```

### Statistical Tests

Automatic p-value calculation with configurable tests:

```{r tests}
# Default tests (t-test for continuous, chi-square for categorical)
table1(arm ~ age + sex, data = data, test = TRUE)

# Specify test types
table1(arm ~ age + sex, data = data,
       continuous_test = "kruskal",
       categorical_test = "fisher")
```

Available tests:

- Continuous: `"ttest"`, `"anova"`, `"welch"`, `"kruskal"`
- Categorical: `"chisq"`, `"fisher"`

### Numeric Summary Options

```{r summaries}
# Mean (SD) - default
table1(arm ~ age, data = data, numeric_summary = "mean_sd")

# Median [IQR]
table1(arm ~ age, data = data, numeric_summary = "median_iqr")

# Mean (95% CI)
table1(arm ~ age, data = data, numeric_summary = "mean_ci")

# Median (range)
table1(arm ~ age, data = data, numeric_summary = "median_range")
```

### Stratified Analysis

Multi-center or multi-site stratification:

```{r stratified}
table1(arm ~ age + sex, data = data, strata = "site")
```

### Missing Data

Display missing value counts:

```{r missing}
table1(arm ~ age + sex, data = data, missing = TRUE)
```

## Output Formats

### Console Output

```{r console}
tbl <- table1(arm ~ age + sex, data = data)
print(tbl)
```
### HTML Output

```{r html}
tbl <- table1(arm ~ age + sex, data = data)
render_html(tbl)
```

### LaTeX Output

```{r latex}
tbl <- table1(arm ~ age + sex, data = data)
render_latex(tbl)
```

## Function Reference

| Function | Purpose |
|:---------|:--------|
| `table1()` | Create Table 1 blueprint |
| `print()` | Console output |
| `render_html()` | HTML output |
| `render_latex()` | LaTeX output |
| `list_available_themes()` | Show available themes |

## Quick Reference

### Formula Syntax

```
grouping_variable ~ var1 + var2 + var3
```

### Key Parameters

| Parameter | Description | Default |
|:----------|:------------|:--------|
| `theme` | Journal theme | "minimal" |
| `test` | Include p-values | FALSE |
| `missing` | Show missing counts | FALSE |
| `numeric_summary` | Summary type | "mean_sd" |
| `strata` | Stratification variable | NULL |

## Next Steps

- `vignette("theming_system")` - Medical journal themes
- `vignette("customizing_statistics")` - Custom summary functions
- `vignette("stratified_examples")` - Multi-center analyses
- `vignette("dataset_examples")` - Complete examples
