---
title: "Quickstart Guide: zztable1"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Quickstart Guide: zztable1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  results = 'asis'
)

library(htmltools)

source("../R/table1.R")
source("../R/blueprint.R")
source("../R/validation_consolidated.R")
source("../R/dimensions.R")
source("../R/cells.R")
source("../R/themes.R")
source("../R/rendering.R")
source("../R/utils.R")
```

```{r theme-css, results='asis', echo=FALSE}
theme_css <- generate_theme_css()
cat("<style>\n")
cat(theme_css)
cat("\n</style>")
```

```{r helper-functions, include=FALSE}
`%||%` <- function(x, y) if (is.null(x)) y else x

create_table <- function(formula, data, ...) {
  args <- list(...)
  theme_arg <- args$theme %||% "console"

  if (is.character(theme_arg)) {
    theme <- get_theme(theme_arg)
  } else if (is.list(theme_arg)) {
    theme <- theme_arg
  } else {
    theme <- get_theme("console")
  }

  bp <- table1(formula = formula, data = data, ...)

  if (knitr::is_latex_output()) {
    output <- render_latex(bp, theme)
  } else if (knitr::is_html_output()) {
    output <- render_html(bp, theme)
  } else {
    output <- render_console(bp, theme)
  }

  knitr::asis_output(paste(output, collapse = "\n"))
}
```

## Overview

**zztable1** is a next-generation R package for creating
publication-quality summary tables (Table 1) with a flexible
blueprint-based architecture. The package features lazy evaluation,
sparse storage, and medical journal themes (NEJM, Lancet, JAMA).

## Installation

```{r install, eval=FALSE}
devtools::install_github("rgt47/zztable1")
library(zztable1)
```

## Sample Data

All examples below use a small simulated clinical trial dataset:

```{r data-setup}
set.seed(42)
n <- 120

trial <- data.frame(
  arm = factor(
    sample(c("Placebo", "Treatment"), n, replace = TRUE),
    levels = c("Placebo", "Treatment")
  ),
  age = round(rnorm(n, 62, 14)),
  sex = factor(
    sample(c("Male", "Female"), n, replace = TRUE)
  ),
  bmi = round(rnorm(n, 27, 5), 1),
  diabetes = factor(
    sample(c("No", "Yes"), n, replace = TRUE,
           prob = c(0.7, 0.3))
  ),
  site = factor(
    sample(paste("Site", LETTERS[1:3]), n,
           replace = TRUE)
  )
)

trial$bmi[sample(seq_len(n), 5)] <- NA
```

## Basic Usage

The formula interface: `grouping_variable ~ var1 + var2 + ...`

```{r basic}
create_table(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  theme = "nejm"
)
```

## Adding P-values and Totals

```{r pvalue-totals}
create_table(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  totals = TRUE,
  theme = "nejm"
)
```

## Medical Journal Themes

### NEJM

```{r theme-nejm}
create_table(
  arm ~ age + sex + bmi,
  data = trial,
  pvalue = TRUE,
  theme = "nejm"
)
```

### Lancet

```{r theme-lancet}
create_table(
  arm ~ age + sex + bmi,
  data = trial,
  pvalue = TRUE,
  theme = "lancet"
)
```

### JAMA

```{r theme-jama}
create_table(
  arm ~ age + sex + bmi,
  data = trial,
  pvalue = TRUE,
  theme = "jama"
)
```

## Numeric Summary Options

```{r summary-mean-sd}
cat("### Mean (SD) -- default\n")
create_table(
  arm ~ age + bmi, data = trial,
  numeric_summary = "mean_sd", theme = "console"
)
```

```{r summary-median-iqr}
cat("\n### Median [IQR]\n")
create_table(
  arm ~ age + bmi, data = trial,
  numeric_summary = "median_iqr", theme = "console"
)
```

```{r summary-mean-ci}
cat("\n### Mean (95% CI)\n")
create_table(
  arm ~ age + bmi, data = trial,
  numeric_summary = "mean_ci", theme = "console"
)
```

## Statistical Tests

Default: t-test for continuous, Fisher's exact for categorical.

```{r tests}
create_table(
  arm ~ age + sex + diabetes,
  data = trial,
  pvalue = TRUE,
  continuous_test = "kruskal",
  categorical_test = "chisq",
  theme = "console"
)
```

Available tests:

- Continuous: `"ttest"`, `"anova"`, `"welch"`, `"kruskal"`
- Categorical: `"chisq"`, `"fisher"`

## Stratified Analysis

```{r stratified}
create_table(
  arm ~ age + sex + bmi,
  data = trial,
  strata = "site",
  pvalue = TRUE,
  theme = "lancet"
)
```

## Missing Data

```{r missing-data}
create_table(
  arm ~ age + sex + bmi,
  data = trial,
  missing = TRUE,
  pvalue = TRUE,
  theme = "jama"
)
```

## Available Themes

```{r list-themes}
cat(paste(list_available_themes(), collapse = ", "), "\n")
```

## Function Reference

| Function | Purpose |
|:---------|:--------|
| `table1()` | Create Table 1 blueprint |
| `print()` | Console output |
| `render_console()` | Rendered console output |
| `render_html()` | HTML output |
| `render_latex()` | LaTeX output |
| `display_table()` | Auto-detect format and display |
| `list_available_themes()` | Show available themes |
| `create_custom_theme()` | Create custom theme |

## Key Parameters

| Parameter | Description | Default |
|:----------|:------------|:--------|
| `theme` | Journal theme | "console" |
| `pvalue` | Include p-values | TRUE |
| `totals` | Include totals column | FALSE |
| `missing` | Show missing counts | FALSE |
| `numeric_summary` | Summary type | "mean_sd" |
| `strata` | Stratification variable | NULL |
| `continuous_test` | Test for continuous vars | "ttest" |
| `categorical_test` | Test for categorical vars | "fisher" |

## Formula Syntax

```
grouping_variable ~ var1 + var2 + var3
```

## Next Steps

- `vignette("theming_system")` -- Medical journal themes
- `vignette("customizing_statistics")` -- Custom summary functions
- `vignette("stratified_examples")` -- Multi-center analyses
- `vignette("dataset_examples")` -- Complete examples
