---
title: "Column Spanners and Summary Rows"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
    extra_dependencies:
      - booktabs
      - colortbl
      - xcolor
      - threeparttable
vignette: >
  %\VignetteIndexEntry{Column Spanners and Summary Rows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  results = 'asis'
)

library(htmltools)

library(zztable1)
```

```{r theme-css, results='asis', echo=FALSE}
if (knitr::is_html_output()) {
  theme_css <- generate_theme_css()
  cat("<style>\n")
  cat(theme_css)
  cat("\n</style>")
}
```

```{r helper, include=FALSE}
`%||%` <- function(x, y) if (is.null(x)) y else x

render_bp <- function(bp, theme_name = "nejm") {
  theme <- get_theme(theme_name)
  if (knitr::is_latex_output()) {
    output <- render_latex(bp, theme)
  } else if (knitr::is_html_output()) {
    output <- render_html(bp, theme)
  } else {
    output <- render_console(bp, theme)
  }
  knitr::asis_output(paste(output, collapse = "\n"))
}
```

## Overview

This vignette demonstrates two structural extensions to the
zztable1 blueprint framework:

- **Column spanners** group columns under hierarchical headers
- **Summary rows** insert computed subtotals at group boundaries

Both features are applied post-hoc to an existing blueprint
using a pipeline-style API.

## Sample Data

```{r data-setup}
set.seed(42)
n <- 120

trial <- data.frame(
  arm = factor(
    sample(c("Placebo", "Treatment"), n, replace = TRUE),
    levels = c("Placebo", "Treatment")
  ),
  age = round(rnorm(n, 62, 14)),
  sex = factor(
    sample(c("Male", "Female"), n, replace = TRUE)
  ),
  bmi = round(rnorm(n, 27, 5), 1),
  diabetes = factor(
    sample(c("No", "Yes"), n, replace = TRUE,
           prob = c(0.7, 0.3))
  )
)
```

## Column Spanners

### Basic spanner

A single-level spanner groups the treatment arm columns under
a shared label.

```{r basic-spanner}
bp <- table1(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  theme = "nejm"
)

add_spanner(bp, label = "Treatment Arms",
            columns = c(2L, 3L))

render_bp(bp)
```

### Multiple spanners

Two separate spanners, each covering different column ranges.

```{r multi-spanner}
bp2 <- table1(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  totals = TRUE,
  theme = "lancet"
)

add_spanner(bp2, label = "By Arm",
            columns = c(2L, 3L), id = "arms")
add_spanner(bp2, label = "Overall",
            columns = 4L, id = "overall")

render_bp(bp2, "lancet")
```

### Nested (hierarchical) spanners

A parent spanner groups two child spanners, producing a
two-level column hierarchy.

```{r nested-spanner}
bp3 <- table1(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  totals = TRUE,
  theme = "jama"
)

add_spanner(bp3, label = "By Arm",
            columns = c(2L, 3L), id = "by_arm")
add_spanner(bp3, label = "Overall",
            columns = 4L, id = "overall")
add_spanner(bp3, label = "Patient Characteristics",
            spanners = c("by_arm", "overall"))

render_bp(bp3, "jama")
```

## Summary Rows

### Per-group summaries

Summary rows are computed over the numeric values in each
variable group and inserted at the group boundary.

```{r group-summaries}
bp4 <- table1(
  arm ~ age + sex + bmi,
  data = trial,
  pvalue = FALSE,
  theme = "nejm"
)

add_summary_rows(bp4,
  fns = list(
    "Group n" = function(x) sum(!is.na(x))
  ),
  side = "bottom"
)

render_bp(bp4)
```

### Grand summary

A grand summary row aggregates across the entire table.

```{r grand-summary}
bp5 <- table1(
  arm ~ age + bmi,
  data = trial,
  pvalue = TRUE,
  theme = "lancet"
)

add_grand_summary_rows(bp5,
  fns = list(
    "Overall Mean" = function(x) round(mean(x, na.rm = TRUE), 1)
  ),
  side = "bottom"
)

render_bp(bp5, "lancet")
```

### Combined: subtotals and grand total

Per-group counts and a grand total in the same table.

```{r combined-summaries}
bp6 <- table1(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  totals = TRUE,
  theme = "jama"
)

add_summary_rows(bp6,
  fns = list(
    "Subtotal n" = function(x) sum(!is.na(x))
  ),
  side = "bottom"
)

add_grand_summary_rows(bp6,
  fns = list(
    "Grand Total n" = function(x) sum(!is.na(x))
  ),
  side = "bottom"
)

render_bp(bp6, "jama")
```

## Both Features Together

Spanners and summary rows on the same blueprint.

```{r both-features}
bp7 <- table1(
  arm ~ age + sex + bmi + diabetes,
  data = trial,
  pvalue = TRUE,
  totals = TRUE,
  theme = "nejm"
)

add_spanner(bp7, label = "Randomised Arms",
            columns = c(2L, 3L), id = "arms")
add_spanner(bp7, label = "All",
            columns = 4L, id = "all")
add_spanner(bp7, label = "Participants",
            spanners = c("arms", "all"))

add_summary_rows(bp7,
  fns = list(
    "n obs" = function(x) sum(!is.na(x))
  ),
  side = "bottom"
)

add_grand_summary_rows(bp7,
  fns = list(
    "Total n" = function(x) sum(!is.na(x))
  ),
  side = "bottom"
)

render_bp(bp7)
```

## API Reference

| Function | Purpose |
|:---------|:--------|
| `add_spanner()` | Add a column spanner to a blueprint |
| `add_spanner_delim()` | Auto-generate spanners from column name delimiters |
| `add_summary_rows()` | Add per-group summary rows |
| `add_grand_summary_rows()` | Add table-wide grand summary rows |

### `add_spanner()` Parameters

| Parameter | Description |
|:----------|:------------|
| `label` | Display text for the spanner header |
| `columns` | Column indices or names to group |
| `id` | Unique identifier (for nesting) |
| `spanners` | Child spanner ids (for nested spanners) |
| `level` | Explicit level override |

### `add_summary_rows()` Parameters

| Parameter | Description |
|:----------|:------------|
| `fns` | Named list of aggregation functions |
| `columns` | Columns to summarise (default: all data columns) |
| `side` | `"bottom"` or `"top"` of each group |
| `groups` | Restrict to specific groups |
| `fmt_fn` | Custom formatting function |
