---
title: "Stratified Table 1 Analysis Examples"
author: "zztable1_nextgen"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    includes:
      in_header: preamble.tex
    extra_dependencies:
      - xcolor
      - colortbl
      - array
      - booktabs
      - longtable
      - float
      - threeparttable
  html_document:
    toc: true
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Stratified Table 1 Analysis Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  results = 'asis'
)

# Load required libraries
library(htmltools)

# Source all required files
source("../R/table1.R")
source("../R/blueprint.R")
source("../R/validation_consolidated.R") 
source("../R/dimensions.R")
source("../R/cells.R")
source("../R/themes.R")
source("../R/rendering.R")
source("../R/utils.R")
```

```{r theme-css, results='asis', echo=FALSE}
# Generate and inject CSS for theme styling
theme_css <- generate_theme_css()
cat("<style>\n")
cat(theme_css)
cat("\n</style>")
```

```{r helper-functions, include=FALSE}
# Helper function for null coalescing
`%||%` <- function(x, y) if (is.null(x)) y else x

# Helper function to create tables with appropriate format
create_table <- function(formula, data, strata = NULL, ...) {
  # Determine output format
  if (knitr::is_latex_output()) {
    layout <- "latex"
    format_type <- "latex"
  } else {
    layout <- "html"
    format_type <- "html"
  }
  
  # Extract theme from arguments
  args <- list(...)
  theme_arg <- args$theme %||% "console"
  
  # Handle both theme names and custom theme objects
  if (is.character(theme_arg)) {
    theme <- get_theme(theme_arg)
  } else if (is.list(theme_arg)) {
    theme <- theme_arg
  } else {
    theme <- get_theme("console")
  }
  
  bp <- table1(formula = formula, data = data, strata = strata, ...)
  
  # Get the rendered output
  output <- switch(format_type,
    "html" = render_html(bp, theme),
    "latex" = render_latex(bp, theme),
    render_console(bp, theme)
  )
  
  if (format_type == "html") {
    return(knitr::asis_output(paste(output, collapse = "\n")))
  } else if (format_type == "latex") {
    return(knitr::asis_output(paste(output, collapse = "\n")))
  } else {
    cat(paste(output, collapse = "\n"))
    cat("\n")
  }
}
```

# Introduction

Stratified analysis is a crucial component of epidemiological and clinical research. This vignette demonstrates how to create stratified Table 1 analyses using the `zztable1_nextgen` package. Stratified tables allow you to examine patterns within subgroups of your data, revealing important differences that might be obscured in overall analyses.

# Clinical Trial Data Example

Let's create a comprehensive clinical trial dataset that includes multiple potential stratification variables.

```{r clinical-data-setup}
# Create a realistic multi-center clinical trial dataset
set.seed(42)
n <- 300

clinical_data <- data.frame(
  # Primary treatment variable
  treatment = factor(
    sample(c("Placebo", "Low Dose", "High Dose"), n, replace = TRUE, 
           prob = c(0.4, 0.3, 0.3)),
    levels = c("Placebo", "Low Dose", "High Dose")
  ),
  
  # Potential stratification variables
  site = factor(sample(paste("Site", LETTERS[1:4]), n, replace = TRUE)),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.55, 0.45))),
  age_group = factor(
    sample(c("18-44", "45-64", "65+"), n, replace = TRUE, prob = c(0.3, 0.4, 0.3)),
    levels = c("18-44", "45-64", "65+")
  ),
  disease_severity = factor(
    sample(c("Mild", "Moderate", "Severe"), n, replace = TRUE, prob = c(0.4, 0.4, 0.2)),
    levels = c("Mild", "Moderate", "Severe")
  ),
  
  # Baseline characteristics
  age = round(rnorm(n, 58, 15)),
  bmi = round(rnorm(n, 26.5, 4.2), 1),
  systolic_bp = round(rnorm(n, 135, 18)),
  
  # Comorbidities
  diabetes = factor(sample(c("No", "Yes"), n, replace = TRUE, prob = c(0.75, 0.25))),
  hypertension = factor(sample(c("No", "Yes"), n, replace = TRUE, prob = c(0.65, 0.35))),
  
  # Lab values
  hemoglobin = round(rnorm(n, 13.2, 1.8), 1),
  creatinine = round(rnorm(n, 1.1, 0.3), 2)
)

# Add some realistic missing values
clinical_data$bmi[sample(1:n, 8)] <- NA
clinical_data$hemoglobin[sample(1:n, 5)] <- NA
clinical_data$creatinine[sample(1:n, 3)] <- NA

# Show dataset structure
str(clinical_data)
head(clinical_data, 10)
```

# Stratified Analysis Examples

## Example 1: Stratified by Study Site

Understanding how baseline characteristics vary across different study sites is crucial for multi-center trials.

```{r site-stratified, results='asis'}
create_table(
  treatment ~ age + sex + bmi + diabetes + systolic_bp,
  data = clinical_data,
  strata = "site",
  theme = "nejm",
  pvalue = TRUE,
  totals = TRUE
)
```

**Key Observations:**
- Each study site shows as a separate table section
- Within-site treatment group comparisons
- Helps identify site-specific recruitment patterns
- Essential for assessing treatment balance across sites

## Example 2: Stratified by Sex

Sex-stratified analysis is important for understanding treatment effects and baseline differences between male and female participants.

```{r sex-stratified, results='asis'}
create_table(
  treatment ~ age + age_group + bmi + diabetes + hypertension + hemoglobin,
  data = clinical_data,
  strata = "sex", 
  theme = "lancet",
  pvalue = TRUE,
  totals = TRUE
)
```

**Key Observations:**
- Separate baseline characteristics for males and females
- Age and BMI distributions may differ by sex
- Hemoglobin levels typically differ between sexes
- Treatment allocation balance within each sex

## Example 3: Stratified by Disease Severity

Disease severity stratification helps understand how patient characteristics and treatment allocation vary by baseline disease status.

```{r severity-stratified, results='asis'}
create_table(
  treatment ~ age + sex + bmi + systolic_bp + diabetes + hypertension + creatinine,
  data = clinical_data,
  strata = "disease_severity",
  theme = "jama",
  pvalue = TRUE,
  totals = TRUE
)
```

**Key Observations:**
- Baseline characteristics across mild, moderate, and severe disease
- Treatment allocation may vary by severity
- Comorbidity prevalence often increases with disease severity
- Important for stratified randomization assessment

## Example 4: Stratified by Age Group

Age-group stratified analysis reveals how baseline characteristics and treatment allocation vary across different age ranges.

```{r age-group-stratified, results='asis'}
create_table(
  treatment ~ sex + bmi + systolic_bp + diabetes + hypertension + hemoglobin + creatinine,
  data = clinical_data,
  strata = "age_group",
  theme = "nejm",
  pvalue = TRUE,
  totals = TRUE
)
```

**Key Observations:**
- Younger vs middle-aged vs older participants
- Comorbidity prevalence increases with age
- Lab values may vary by age group
- Treatment allocation balance across age groups

# Advanced Stratified Analysis

## Multiple Variables with Missing Data

Let's examine how stratified analysis handles missing data and multiple variable types.

```{r missing-data-stratified, results='asis'}
create_table(
  treatment ~ age + bmi + hemoglobin + creatinine + diabetes + hypertension,
  data = clinical_data,
  strata = "sex",
  theme = "lancet", 
  missing = TRUE,  # Show missing value patterns
  pvalue = TRUE,
  totals = TRUE
)
```

## Site and Sex Combined Analysis

For comprehensive analysis, we might want to examine the interaction of multiple stratification factors.

```{r site-sex-overview}
# Create a combined stratification variable for demonstration
clinical_data$site_sex <- interaction(clinical_data$site, clinical_data$sex, sep = " - ")

# Show the distribution
table(clinical_data$site_sex, clinical_data$treatment)
```

```{r site-sex-stratified, results='asis'}
create_table(
  treatment ~ age + bmi + diabetes + systolic_bp,
  data = clinical_data,
  strata = "site_sex",
  theme = "jama",
  pvalue = TRUE
)
```

# Comparative Analysis

## Before and After Stratification

Let's compare an overall analysis with a stratified analysis to see how stratification reveals important patterns.

### Overall Analysis (Non-stratified)

```{r overall-analysis, results='asis'}
create_table(
  treatment ~ age + sex + bmi + diabetes + hypertension + systolic_bp,
  data = clinical_data,
  theme = "console",
  pvalue = TRUE,
  totals = TRUE
)
```

### Stratified by Disease Severity

```{r comparison-stratified, results='asis'}
create_table(
  treatment ~ age + sex + bmi + diabetes + hypertension + systolic_bp,
  data = clinical_data,
  strata = "disease_severity",
  theme = "console", 
  pvalue = TRUE,
  totals = TRUE
)
```

**Key Differences:**
- Overall analysis may mask important subgroup differences
- Stratified analysis reveals severity-specific patterns
- P-values may differ when accounting for stratification
- Treatment balance assessment within severity levels

# Summary and Best Practices

## When to Use Stratified Analysis

1. **Multi-center Studies**: Always stratify by study site
2. **Sex Differences**: Important for most clinical studies
3. **Age Groups**: Especially relevant for studies spanning wide age ranges
4. **Disease Severity**: Critical for understanding baseline risk
5. **Geographic Regions**: For studies spanning different populations

## Interpretation Guidelines

1. **Sample Sizes**: Check adequate sample sizes within strata
2. **Missing Data**: Consider missing data patterns within strata
3. **P-values**: Interpret within-strata comparisons carefully
4. **Clinical Relevance**: Focus on clinically meaningful differences
5. **Multiple Comparisons**: Consider adjustment for multiple testing

## Available Stratification Variables in This Dataset

```{r strata-summary, echo=FALSE}
strata_vars <- data.frame(
  Variable = c("site", "sex", "age_group", "disease_severity", "diabetes", "hypertension"),
  Description = c(
    "Study site (A, B, C, D)",
    "Participant sex (Male, Female)", 
    "Age groups (18-44, 45-64, 65+)",
    "Disease severity (Mild, Moderate, Severe)",
    "Diabetes status (No, Yes)",
    "Hypertension status (No, Yes)"
  ),
  `Use Case` = c(
    "Multi-center trial balance",
    "Sex-specific effects",
    "Age-related patterns", 
    "Baseline risk stratification",
    "Comorbidity analysis",
    "Cardiovascular risk factors"
  ),
  check.names = FALSE
)

knitr::kable(strata_vars, caption = "Available Stratification Variables")
```

The stratified analysis capabilities of `zztable1_nextgen` provide powerful tools for understanding complex clinical trial data. By examining baseline characteristics within meaningful subgroups, researchers can better assess treatment allocation, identify potential confounders, and plan appropriate statistical analyses.