---
title: "zztable1: Creating Publication-Ready Summary Tables for Clinical Trials"
author: "Ronald (Ryy) G. Thomas"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    highlight: tango
vignette: >
  %\VignetteIndexEntry{zztable1: Creating Publication-Ready Summary Tables for Clinical Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = 'center',
  comment = "#>"
)
```

# Introduction

The `zztable1` package provides tools for creating publication-ready summary tables for clinical trial data. It's particularly focused on creating "Table 1" - the baseline characteristics table that is a standard component of clinical trial publications. This vignette demonstrates the key functionalities of the package with worked examples.

# Installation

You can install the development version of `zztable1` from GitHub with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("rgt47/zztable1")
```

```{r}
# Load required packages
library(zztable1)
```

# Creating Sample Dataset

For this vignette, we'll create a sample clinical trial dataset to demonstrate the package:

```{r}
# Create a sample dataset
set.seed(123)
trial_data <- data.frame(
  arm = factor(rep(c("Treatment", "Placebo"), each = 50)),
  age = rnorm(100, mean = 45, sd = 15),
  sex = factor(sample(c("Male", "Female"), 100, replace = TRUE)),
  bmi = rnorm(100, mean = 26, sd = 5),
  diabetes = factor(sample(c("Yes", "No"), 100, replace = TRUE, prob = c(0.3, 0.7))),
  site = factor(sample(c("Site1", "Site2", "Site3"), 100, replace = TRUE))
)

# Add some missing values for demonstration
trial_data$age[sample(1:100, 5)] <- NA
trial_data$bmi[sample(1:100, 8)] <- NA

# View the first few rows of the data
head(trial_data)
```

# Basic Usage

## Creating a Simple Table 1

The main function in the package is `table1()`, which takes a formula specifying the grouping variable and the variables to summarize:

```{r}
# Basic table with treatment arm as the grouping variable
basic_table <- table1(form = arm ~ age + sex + bmi + diabetes, data = trial_data)
print(basic_table)
```

In this example:
- `arm` is the grouping variable (treatment groups)
- `age`, `sex`, `bmi`, and `diabetes` are the variables to be summarized
- The formula syntax is similar to that used in modeling functions in R

## Adding Totals

You can add a "Total" column to see overall statistics:

```{r}
# Table with totals
totals_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                      data = trial_data, 
                      totals = TRUE)
print(totals_table)
```

## Showing Missing Values

To include counts of missing values for each variable:

```{r}
# Table with missing value counts
missing_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                       data = trial_data, 
                       missing = TRUE)
print(missing_table)
```

For numeric variables, this adds a "valid (missing)" row showing the count of non-missing values and the count of missing values.

## Including Group Sizes

To show the number of subjects in each group:

```{r}
# Table with group sizes
size_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                    data = trial_data, 
                    size = TRUE)
print(size_table)
```

## Disabling P-values

By default, the table includes p-values for comparing groups. You can disable this:

```{r}
# Table without p-values
nopvalue_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                        data = trial_data, 
                        pvalue = FALSE)
print(nopvalue_table)
```

# Advanced Usage

## Stratification

One powerful feature of `zztable1` is the ability to stratify tables by another variable:

```{r}
# Stratified table by study site
stratified_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                         data = trial_data, 
                         strata = "site")
print(stratified_table)
```

This creates separate tables for each level of the stratification variable, allowing you to see how group comparisons vary across different strata.

## Combining Options

You can combine multiple options to create customized tables:

```{r}
# Comprehensive table with multiple options
comprehensive_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                            data = trial_data, 
                            totals = TRUE, 
                            missing = TRUE, 
                            size = TRUE)
print(comprehensive_table)
```

## Creating Tables Without a Grouping Variable

If you just want summary statistics without group comparisons:

```{r}
# Table without group comparisons
overall_table <- table1(form = ~ age + sex + bmi + diabetes, 
                      data = trial_data, 
                      totals = TRUE,
                      pvalue = FALSE)
print(overall_table)
```

Note: When using a formula without a grouping variable, you must set `totals = TRUE` and `pvalue = FALSE`.

# LaTeX Output

The package includes functionality to export tables in LaTeX format, which is particularly useful for publications:

```{r, eval=FALSE}
# Create a table
my_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                  data = trial_data, 
                  totals = TRUE)

# Export to LaTeX with default NEJM styling
latex_table <- latex(my_table, 
                    digits = 2, 
                    fname = "my_clinical_trial_table")
```

This creates a LaTeX file with the table, styled according to the New England Journal of Medicine (NEJM) formatting guidelines.

## Customizing LaTeX Output

You can customize the appearance of your tables by defining your own theme:

```{r, eval=FALSE}
# Define a custom theme
my_theme <- list(
  foreground = c("black", "darkblue", "black", "darkblue", "darkred"),
  background = c("#f0f0f0", "white", "#e0e0e0", "white", "#fff0f0")
)

# Export to LaTeX with custom styling
custom_latex_table <- latex(my_table, 
                          digits = 2, 
                          fname = "custom_styled_table", 
                          theme = my_theme)
```

# Real-World Examples

## Example 1: Analyzing Clinical Trial Baseline Characteristics

Let's analyze a more realistic clinical trial dataset with multiple treatment arms:

```{r}
# Create a more detailed sample dataset
set.seed(456)
n <- 300
detailed_trial <- data.frame(
  treatment = factor(sample(c("Drug A", "Drug B", "Placebo"), n, replace = TRUE)),
  age = rnorm(n, mean = 62, sd = 12),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.55, 0.45))),
  race = factor(sample(c("White", "Black", "Asian", "Other"), n, replace = TRUE, 
                      prob = c(0.70, 0.15, 0.10, 0.05))),
  bmi = rnorm(n, mean = 28, sd = 6),
  diabetes = factor(sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.25, 0.75))),
  hypertension = factor(sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.40, 0.60))),
  center = factor(sample(1:10, n, replace = TRUE))
)

# Create a comprehensive baseline characteristics table
baseline_table <- table1(form = treatment ~ age + sex + race + bmi + 
                               diabetes + hypertension, 
                       data = detailed_trial, 
                       totals = TRUE, 
                       size = TRUE, 
                       missing = FALSE)
print(baseline_table)
```

## Example 2: Stratified Analysis by Study Center

For multi-center trials, it's often important to examine baseline characteristics by center:

```{r eval =FALSE}
# Analyze baseline characteristics stratified by study center
# We'll use a subset of centers for clarity
center_subset <- detailed_trial[detailed_trial$center %in% c(1, 2, 3), ]

center_stratified <- table1(form = treatment ~ age + sex + bmi + diabetes + hypertension,
                          data = center_subset,
                          strata = "center",
                          totals = TRUE)
print(center_stratified)
```

## Example 3: Analyzing Outcomes Without Group Comparisons

Sometimes you just want to describe outcomes without formal comparisons:

```{r}
# Create some outcome data
detailed_trial$adverse_events <- rpois(n, lambda = 1.5)
detailed_trial$completed_study <- factor(sample(c("Yes", "No"), n, replace = TRUE, 
                                              prob = c(0.85, 0.15)))
detailed_trial$response <- factor(sample(c("Complete", "Partial", "None"), n, replace = TRUE, 
                                       prob = c(0.30, 0.40, 0.30)))

# Summarize outcomes without group comparisons
outcome_table <- table1(form = ~ adverse_events + completed_study + response,
                      data = detailed_trial,
                      totals = TRUE,
                      pvalue = FALSE)
print(outcome_table)
```

## Example 4: Tables for Safety Analysis

Safety analysis is a common task in clinical trials. Here's how to create a table for adverse events by severity:

```{r}
# Create adverse event data
# Create a simpler adverse event data
set.seed(789)
ae_data <- data.frame(
  treatment = factor(rep(c("Drug A", "Drug B", "Placebo"), each = 30)),  # Reduced from 100 each
  headache = factor(sample(c("None", "Mild", "Severe"), 90, replace = TRUE, 
                         prob = c(0.6, 0.3, 0.1))),  # Reduced severity levels
  nausea = factor(sample(c("None", "Mild", "Severe"), 90, replace = TRUE,
                       prob = c(0.7, 0.2, 0.1))),
  dizziness = factor(sample(c("None", "Present"), 90, replace = TRUE,  # Binary factor
                          prob = c(0.8, 0.2)))
)

# Create safety table
safety_table <- table1(form = treatment ~ headache + nausea + dizziness,
                     data = ae_data,
                     totals = TRUE,
                     size = TRUE)
print(safety_table)
```

## Example 5: Tables with Laboratory Values 

Laboratory values often have their own format requirements:

```{r}
# Create lab data
set.seed(101)
lab_data <- data.frame(
  treatment = factor(rep(c("Drug A", "Drug B", "Placebo"), each = 100)),
  hemoglobin = rnorm(300, mean = 14, sd = 1.5),
  wbc = rnorm(300, mean = 7.5, sd = 2),
  platelets = rnorm(300, mean = 250, sd = 50),
  creatinine = rnorm(300, mean = 0.9, sd = 0.2),
  alt = rnorm(300, mean = 30, sd = 15),
  ast = rnorm(300, mean = 25, sd = 10)
)

# Add some out-of-range values
lab_data$alt[sample(1:300, 15)] <- lab_data$alt[sample(1:300, 15)] * 3
lab_data$ast[sample(1:300, 10)] <- lab_data$ast[sample(1:300, 10)] * 2.5

# Create lab value table
lab_table <- table1(form = treatment ~ hemoglobin + wbc + platelets + creatinine + alt + ast,
                  data = lab_data,
                  totals = TRUE,
                  size = TRUE)
print(lab_table)
```

# Tips and Best Practices

1. **Variable types matter**: The way variables are summarized depends on their R data type:
   - Numeric variables are summarized with mean and standard deviation
   - Factors are summarized with counts and percentages
   - Make sure your variables have the appropriate type

2. **Handle missing data appropriately**: Use the `missing = TRUE` option to see counts of missing values, which is important for assessing data quality

3. **Consider p-value interpretation**: Remember that p-values should be interpreted in the context of multiple testing, especially when comparing many variables

4. **Customizing LaTeX output**: For publications, spend time customizing the LaTeX output to meet journal requirements

5. **Export to other formats**: While LaTeX output is provided natively, you can also use functions from packages like `knitr` to convert the tables to other formats like HTML or Word

# Integration with Other Packages

The `zztable1` package works well with other R packages:

```{r, eval=FALSE}
# Create a table
my_table <- table1(form = arm ~ age + sex + bmi + diabetes, 
                  data = trial_data, 
                  totals = TRUE)

# Export to HTML using kableExtra
library(kableExtra)
html_table <- my_table |>
  dplyr::select(-code) |>  # Remove the code column
  kbl(format = "html") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# View in browser
if (interactive()) {
  print(html_table)
}
```

# Summary

The `zztable1` package provides a streamlined way to create publication-ready summary tables for clinical trial data. With its formula-based interface and flexible options for customization, it can handle a wide range of table requirements for clinical research.

Key features include:
- Simple formula interface for specifying variables
- Automatic calculation of appropriate summary statistics based on variable type
- Calculation of p-values for group comparisons
- Stratification by categorical variables
- LaTeX output with customizable styling
- Support for missing value reporting

These capabilities make `zztable1` a valuable tool for clinical researchers who need to efficiently create high-quality summary tables for publications, reports, and presentations.

# Session Info

```{r}
sessionInfo()
```
