---
title: "zztable1_nextgen: Advanced Publication-Ready Summary Tables"
author: "zztable1_nextgen Development Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
vignette: >
  %\VignetteIndexEntry{zztable1_nextgen: Advanced Publication-Ready Summary Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  results = 'asis'
)

# Load required libraries
library(htmltools)
library(kableExtra)

# Source all required files
source("../R/table1.R")
source("../R/blueprint.R")
source("../R/validation_consolidated.R")
source("../R/dimensions.R")
source("../R/cells.R")
source("../R/themes.R")
source("../R/rendering.R")
source("../R/utils.R")
```

```{r helper-functions, include=FALSE}
# Helper function for null coalescing
`%||%` <- function(x, y) if (is.null(x)) y else x

# Helper function to create tables with appropriate format
create_table <- function(formula, data, ...) {
  # Extract theme from arguments
  args <- list(...)
  theme_arg <- args$theme %||% "console"

  # Handle both theme names and custom theme objects
  if (is.character(theme_arg)) {
    theme <- get_theme(theme_arg)
  } else if (is.list(theme_arg)) {
    theme <- theme_arg
  } else {
    theme <- get_theme("console")
  }

  bp <- table1(formula = formula, data = data, ...)

  # Detect output format and render appropriately
  if (knitr::is_latex_output()) {
    output <- render_latex(bp, theme)
  } else if (knitr::is_html_output()) {
    output <- render_html(bp, theme)
  } else {
    # Default to console for unknown formats
    output <- render_console(bp, theme)
  }

  return(knitr::asis_output(paste(output, collapse = "\n")))
}
```

# Introduction

The `zztable1_nextgen` package provides a next-generation architecture for
creating publication-ready summary tables (commonly called "Table 1") used
in biomedical research and clinical trials. This vignette demonstrates the
key features and capabilities of the package.

## Key Features

- **Lazy Evaluation Architecture**: Fast blueprint creation with computation
  on demand
- **Journal-Specific Theming**: NEJM, Lancet, JAMA, BMJ formatting styles
- **Advanced Footnote System**: Variable-specific, column-specific, and
  general footnotes with superscript markers
- **Multiple Output Formats**: Console, LaTeX, and HTML output with proper column headers
- **Flexible Statistics**: Built-in and custom summary statistics
- **Stratified Analysis**: Support for subgroup analyses
- **Full Compatibility**: Same interface as original zztable1 package
- **R Markdown Integration**: Automatic format detection for seamless PDF/HTML output

## Installation and Setup

```{r install, eval=FALSE}
# Development version (when available)
# devtools::install_github("user/zztable1_nextgen")

# For this vignette, source the development file
source("zztable1_nextgen.R")
```

# Basic Usage

## Simple Summary Tables

Let's start with a basic example using the mtcars dataset:

```{r basic_example}
# Prepare data
data(mtcars)
mtcars$transmission <- factor(
  ifelse(mtcars$am == 1, "Manual", "Automatic"),
  levels = c("Automatic", "Manual")
)
mtcars$engine_type <- factor(
  ifelse(mtcars$vs == 1, "V-shaped", "Straight"),
  levels = c("Straight", "V-shaped")
)

# Create basic summary table
create_table(transmission ~ mpg + hp + wt, data = mtcars)
```

## Adding Statistical Tests

Include p-values for group comparisons:

```{r with_pvalues}
create_table(transmission ~ mpg + hp + wt,
             data = mtcars,
             pvalue = TRUE)
```

## Including Total Column

Add an overall summary column:

```{r with_totals}
create_table(transmission ~ mpg + hp + wt,
             data = mtcars,
             pvalue = TRUE,
             totals = TRUE)
```

# Advanced Features

## Custom Numeric Summaries

### Built-in Options

The package provides several built-in summary statistics:

```{r numeric_summaries}
# Default: Mean (SD)
cat("Mean (SD) format:\n")
create_table(transmission ~ mpg + hp, data = mtcars)

# Median [IQR]
cat("\nMedian [IQR] format:\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             numeric_summary = "median_iqr")

# Mean +/- SE
cat("\nMean +/- SE format:\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             numeric_summary = "mean_se")
```

### Custom Functions

Create your own summary statistics:

```{r custom_summary}
# Custom function: Median (Min-Max)
custom_summary <- function(x) {
  med <- round(median(x, na.rm = TRUE), 1)
  min_val <- round(min(x, na.rm = TRUE), 1)
  max_val <- round(max(x, na.rm = TRUE), 1)
  paste0(med, " (", min_val, "-", max_val, ")")
}

cat("Custom Median (Min-Max) format:\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             numeric_summary = custom_summary)
```

## Stratified Analysis

Perform subgroup analyses using stratification:

```{r stratified}
# Create stratification variable
mtcars$cylinder_group <- factor(
  ifelse(mtcars$cyl <= 4, "4-cylinder",
  ifelse(mtcars$cyl <= 6, "6-cylinder", "8-cylinder")),
  levels = c("4-cylinder", "6-cylinder", "8-cylinder")
)

# Stratified analysis
create_table(transmission ~ mpg + hp,
             data = mtcars,
             strata = "cylinder_group",
             pvalue = TRUE)
```

# Journal-Specific Theming

## Available Themes

View all available themes:

```{r list_themes}
themes <- list_available_themes()
print(themes)
```

## Theme Comparison

### Default Theme

```{r theme_default}
cat("Default Theme:\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             theme = "default")
```

### NEJM Theme (1 decimal place)

```{r theme_nejm}
cat("NEJM Theme (1 decimal place):\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             theme = "nejm")
```

### JAMA Theme (2 decimal places)

```{r theme_jama}
cat("JAMA Theme (2 decimal places):\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             theme = "jama")
```

### Lancet Theme

```{r theme_lancet}
cat("Lancet Theme:\n")
create_table(transmission ~ mpg + hp, data = mtcars,
             theme = "lancet")
```

# Footnote System

## Variable-Specific Footnotes

Add footnotes to specific variables with superscript markers:

```{r footnotes_variables}
create_table(transmission ~ mpg + hp + wt,
             data = mtcars,
             theme = "nejm",
             footnotes = list(
               variables = list(
                 mpg = "EPA fuel economy rating in miles per gallon",
                 hp = "Gross horsepower measured at crankshaft",
                 wt = "Vehicle weight in thousands of pounds"
               )
             ))
```

## Column-Specific Footnotes

Add footnotes to columns:

```{r footnotes_columns}
create_table(transmission ~ mpg + hp,
             data = mtcars,
             theme = "nejm",
             pvalue = TRUE,
             footnotes = list(
               columns = list(
                 "p.value" = "Two-tailed t-test, alpha = 0.05"
               )
             ))
```

## Comprehensive Footnotes

Combine multiple footnote types:

```{r footnotes_comprehensive}
create_table(transmission ~ mpg + hp,
             data = mtcars,
             theme = "nejm",
             pvalue = TRUE,
             footnotes = list(
               variables = list(
                 mpg = "EPA fuel economy standard",
                 hp = "Gross horsepower"
               ),
               columns = list(
                 "p.value" = "Statistical significance testing"
               ),
               general = list(
                 "Data source: Henderson and Velleman (1981)",
                 "Missing values excluded from analysis"
               )
             ))
```

# Clinical Trial Example

## Simulated Clinical Trial Data

Let's create a more realistic clinical trial example:

```{r clinical_trial_data}
set.seed(123)
n <- 200

# Generate clinical trial data
trial_data <- data.frame(
  patient_id = 1:n,
  treatment = factor(
    sample(c("Placebo", "Drug A", "Drug B"), n, replace = TRUE),
    levels = c("Placebo", "Drug A", "Drug B")
  ),
  age = round(rnorm(n, 65, 12)),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  race = factor(
    sample(c("White", "Black", "Hispanic", "Asian", "Other"), 
           n, replace = TRUE, prob = c(0.6, 0.2, 0.1, 0.08, 0.02)),
    levels = c("White", "Black", "Hispanic", "Asian", "Other")
  ),
  baseline_bmi = round(rnorm(n, 28, 5), 1),
  diabetes = factor(sample(c("No", "Yes"), n, replace = TRUE, prob = c(0.7, 0.3))),
  hypertension = factor(sample(c("No", "Yes"), n, replace = TRUE, prob = c(0.6, 0.4))),
  center = factor(sample(paste("Center", 1:4), n, replace = TRUE))
)

# Preview the data
head(trial_data, 10)
```

## Basic Clinical Table 1

```{r clinical_basic}
create_table(treatment ~ age + sex + race + baseline_bmi +
             diabetes + hypertension,
             data = trial_data,
             theme = "nejm",
             pvalue = TRUE)
```

## With Footnotes and Stratification

```{r clinical_advanced}
create_table(treatment ~ age + sex + race + baseline_bmi +
             diabetes + hypertension,
             data = trial_data,
             strata = "center",
             theme = "nejm",
             pvalue = TRUE,
             footnotes = list(
               variables = list(
                 age = "Age at enrollment (years)",
                 baseline_bmi = "Body mass index at baseline (kg/m²)",
                 diabetes = "Type 2 diabetes mellitus diagnosis",
                 hypertension = "Hypertension diagnosis"
               ),
               columns = list(
                 "p.value" = "ANOVA for continuous, chi-squared for categorical"
               ),
               general = list(
                 "Data are mean (SD) or n (%)",
                 "ITT population (N=200)"
               )
             ))
```

# Different Output Formats

## Console Output (Default)

```{r format_console}
create_table(transmission ~ mpg + hp, data = mtcars, theme = "nejm")
```

## LaTeX Output

```{r format_latex}
bp_latex <- table1(transmission ~ mpg + hp, data = mtcars,
                   layout = "latex", theme = "nejm")

# Note: LaTeX output would contain LaTeX markup
cat("LaTeX theme config:\n")
cat("Font size:", bp_latex$metadata$theme$latex$font_size, "\n")
cat("Packages:", paste(bp_latex$metadata$theme$latex$packages, collapse = ", "), "\n")
```

## HTML Output

```{r format_html}
bp_html <- table1(transmission ~ mpg + hp, data = mtcars,
                  layout = "html", theme = "nejm")

# Note: HTML output would contain HTML markup
cat("HTML theme ready for web display\n")
```

# Performance and Architecture

## Blueprint Architecture

The lazy evaluation approach provides several benefits:

```{r performance_demo}
# Large dataset simulation
large_data <- data.frame(
  group = factor(sample(c("A", "B", "C"), 10000, replace = TRUE)),
  var1 = rnorm(10000),
  var2 = rnorm(10000),
  var3 = rnorm(10000),
  var4 = rnorm(10000),
  var5 = rnorm(10000)
)

# Fast blueprint creation (no computations yet)
system.time({
  bp_large <- table1(group ~ var1 + var2 + var3 + var4 + var5, 
                     data = large_data)
})

# Computations happen only during display
cat("Blueprint created instantly. Computations happen during display.\n")
cat("Blueprint dimensions:", dim(bp_large), "\n")
```

## Memory Efficiency

```{r memory_demo}
# Blueprint object structure
bp_small <- table1(transmission ~ mpg, data = mtcars)

cat("Blueprint components:\n")
cat("- Cells: ", length(bp_small$cells), "\n")
cat("- Dimensions: ", dim(bp_small), "\n")
cat("- Metadata keys: ", names(bp_small$metadata), "\n")
```

# Best Practices

## Recommendations

1. **Choose Appropriate Themes**: Use journal-specific themes for manuscript
   preparation
2. **Add Informative Footnotes**: Explain variables and statistical methods
3. **Use Stratification Wisely**: For meaningful subgroup analyses
4. **Custom Functions**: Create domain-specific summary statistics
5. **Validate Results**: Check statistical assumptions and interpret p-values
   carefully

## Common Patterns

```{r best_practices}
# Standard clinical trial baseline table
create_baseline_table <- function(data, treatment_var, theme = "nejm") {
  formula_str <- paste(treatment_var, "~ .")
  bp <- table1(as.formula(formula_str), 
               data = data,
               theme = theme,
               pvalue = TRUE,
               footnotes = list(
                 general = list(
                   "Data are mean (SD) or n (%)",
                   "P-values from ANOVA or chi-squared test"
                 )
               ))
  return(bp)
}

# Example usage
# bp_standard <- create_baseline_table(trial_data, "treatment")
cat("Utility function created for standardized baseline tables\n")
```

# Troubleshooting

## Common Issues

1. **Missing Variables**: Ensure all formula variables exist in the data
2. **Factor Levels**: Check factor level ordering for expected display
3. **Missing Values**: Use `missing = TRUE` to show missing counts
4. **Theme Application**: Themes affect decimal places and formatting
5. **Large Tables**: Use stratification to break down complex tables

## Error Handling

```{r error_handling, error=TRUE}
# Example of error handling
tryCatch({
  # This will cause an error - variable doesn't exist
  bp_error <- table1(nonexistent_var ~ mpg, data = mtcars)
}, error = function(e) {
  cat("Error caught:", e$message, "\n")
  cat("Solution: Check that all variables in formula exist in data\n")
})
```

# Conclusion

The `zztable1_nextgen` package provides a powerful, flexible system for
creating publication-ready summary tables. Key advantages include:

- **Performance**: Lazy evaluation for fast blueprint creation
- **Flexibility**: Multiple themes, custom statistics, advanced footnotes  
- **Compatibility**: Same interface as original zztable1
- **Publication-Ready**: Journal-specific formatting out of the box

For more information, see the package documentation and function help files.

---

## Session Information

```{r session_info}
sessionInfo()
```